|

### 2019-10-02 analysis/paper/manuscript.Rmd
Address the remainder of Keith's edits/comments

Northern cod (NAFO Divisions 2J3KL) has a rich history of tagging, starting in in 1954 [@taggart1995] and continuing to this day. The tagging and recovery data are captured in a standardized database, with fields typical of most tagging programs. This data base has over 600,000 records as of early 2019, with 2,000-10,000 tags deployed annually in recent decades. The tagging and capture data are used in the current assessment model for this stock [@cadigan2016], but tools to explore this extensive data set were limited, especially from a spatial perspective. To begin to explore and understand this large data set, we built a simple shiny [@chang2018a] application using the mapping package leaflet [@cheng2018], all within the framework of the R programming language [@R] and the RStudio IDE [@Rstudio]. Shiny provides a means for the user to easily interface with the application, including options to quickly and dynamically subset the data (e.g. ranges of release and capture years, specific geographic locations). As the application developed, it was incorporated into shinydashboard [@chang2018b], to take advantage of the number of layout options available in shinydashboard. Given the large number of tags released, often at nearby sites, visualizing the data with static mapping was particularly challenging. The markercluster (https://github.com/Leaflet/Leaflet.markercluster) function available in leaflet was particularly useful as a means to dynamically scale the level of pooling of spatial points (Figure \@ref(fig:map)). This basic mapping tool allowed us to quickly become familiar with the data, identify outliers and incorrect data entries, as well as explore options on how to spatially pool the data for subsequent demographic analysis. Further tabs were added to provide basic summaries of the selected data (Figure \@ref(fig:summary)).

@@

Synthesizing data from multiple sources presents a key challenge to stock assessment. Analyses of different data sources were traditionally carried out independently and the summaries or parameters from these analyses were used in the assessment model. This approach, however, is less than ideal because information may be lost and uncertainty may be unaccounted for when we "do statistics on the statistics" [@link1999; @maunder2013]. Such issues have largely been curtailed in contemporary stock assessments thanks to advances in statistical computing that have facilitated the analysis of all available data, in as raw a form as appropriate, in a single integrated analysis [@maunder2013]. Specifically, statistical modeling tools such as JAGS [@plummer2003], AD Model Builder [@fournier2012] and Template Model Builder [@kristensen2015] allow the construction of a joint likelihood for an array of observations to, in theory, extract as much information as possible about the biological and fishery processes. From a computational perspective, analyses of a variety of large data sets has never been easier. However, from a human perspective, contemporary stock assessment biologists are faced with the challenge of understanding and integrating data from multiple sources into a single model and communicating the methods and results to stakeholders and fisheries managers. Using a recently developed tool for exploring the integrated assessment model for Northern cod as an example, we hope to demonstrate that interactive visualization tools can help meet these challenges.

The Northern cod stock off southern Labrador and eastern Newfoundland is one of the most well studied stocks in eastern Canada, perhaps by virtue of its history. As such, there are multiple monitoring programs that help inform the status of the stock and data from most of these programs have been integrated into a state-space stock assessment model, called NCAM [@cadigan2016]. The model includes information from research vessel autumn trawl surveys (1983-present), Sentinel fishery surveys (1995-present), inshore acoustic surveys (1995-2009), fishery catch-at-age compositions and partial fishery landings (1983-present), and tagging (1983-present). Using a series of observation equations, this TMB based model reduces thousands of historical data points into quantities such as recruitment, spawning stock biomass, fishing mortality and natural mortality. Once the model is fit to the data, the next step is to produce visual representations of the data and model output. The usual approach would be to produce static presentations and documents with a series of figures and tables, but with large amounts of model inputs and outputs, this approach quickly becomes overwhelming for both the analyst and the stakeholders involved. First, it is no longer feasible for the analyst to include and describe every figure and table produced in a single document. Second, it is difficult for stakeholders to efficiently digest the information that has been compressed into a series of static slides or pages. Interactive documents provide a potential solution to this problem as they allow much more information to be contained and accessible on a single screen.

In the pursuit of an easier and more efficient way to communicate results from NCAM, an interactive and self-contained "dashboard" was developed for the 2018 assessment of Northern cod [@dwyer2019]. We used R-based packages [@R] and the RStudio IDE [@Rstudio] to construct a tool for exploring the input and output of NCAM, specifically the flexdashboard [@allaire2017] package to group interactive plotly-based [@sievert2018] visuals into a dynamic document. We also used the crosstalk [@cheng2016] package to link the data displayed across multiple plots. Via R Markdown [@allaire2018], the dashboard is rendered into a self-contained html file that is reproducible, interactive, and easy to update following modifications to the model or the addition of new data.

The NCAM dashboard (Supplement 2) contains a series of pages, the first of which provides terse point-form background on the model (page named "Background"). Subsequent pages provide a series of diagnostic plots for assessing model fits to catch ("Catch"), survey ("RV survey", "SN survey" and "SS survey"; accessed from the "Surveys" drop-down menu), and tagging ("Tagging") data. For instance, the "RV survey" page includes plots of observed and predicted values of mean numbers per tow captured in the research vessel survey (Figure \@ref(fig:rv)). The dashboard also includes pages focused on model estimates such as catchability and selectivity ("Catchability"), stock size and vital rates ("Trends"; Figure \@ref(fig:trends)), and stock productivity ("Productivity"). Finally, some results from a retrospective analyses are included under the "Retro" page, trends from different models are compared under the "Comps" page, details on the projections are accessed from the "Projections" drop-down menu ("Assumptions", "Past projections", "Retro projections", and "Results" pages), and key inputs and outputs are accessed from "Tables" drop-down menu ("Inputs", "Settings", "Outputs" pages). The plots and tables included in the dashboard are similar to those typically presented at assessment meetings and in research documents, however, there are two key benefits of this approach over the *modus operandi* of producing static documents and slides: 1) interactive plots nested in a dashboard permit relatively easy and efficient access to the details as it replaces scrolling through tens, if not hundreds, of pages or slides with mouse-clicks across pages holding data-rich illustrations (i.e. both broader patterns and finer details in the data are accessible via zooming and tooltips); and, 2) the automated nature of the dashboard circumvents the monotonous, time-consuming and error-prone task of copying and pasting figures, tables and values into documents and slides. Both benefits expedite the process of exploring a range of model configurations as the automated output facilitate quick views of standard diagnostics and the interactive plots facilitate detailed explorations and model to model comparisons. The dashboard also makes it easy to share the results with colleagues and stakeholders as it is rendered into a self-contained html file. This allows others to independently scrutinize details of the data and model that are typically only accessible to the analyst. Such access improves the transparency of the stock assessment model which, in turn, leads to richer discussion and scrutiny of the biological and statistical rigor of the model. For instance, visualizations depicting the model assumptions, process errors and confidence intervals around the projections raised important questions on the impact of the assumptions on the projections from the model. These questions were raised during the first assessment meeting in which this tool was used and the ensuing discussion helped the assessment biologists, managers and fisheries representatives in the room gain a deeper understanding of the uncertainties and the patterns in the risk tables [e.g Prob(SSB~2020~ &ge; B~lim~)].

@@

The tools presented here are additions to the growing number of interactive stock assessment tools being developed and applied across the globe (see Appendix A). We  believe that such tools represent an important a step forward in terms of open science and reproducible research in addition to being a major step forward in how stock assessments are presented and critiqued. While these tools may not reveal the entire data-pipeline, it is unlikely that an individual has the time to fully review and recreate a particular analysis in its entirety when most participants are already overcommitted [@banks2011]. In the case of NCAM, considerable statistical experience and expertise is required simply to run the model. However, the dashboard removes many of these difficulties. The html file contains all model inputs, outputs and diagnostics in an open and easy to use interactive document that can be distributed prior to meetings and, as such, participants can assess the results and diagnostics at their leisure. Specific experience with the tools used to generate the results are not required to constructively and critically review of the results presented in the dashboard.

@@

We must acknowledge there are costs to adopting the use of interactive tools. The first and obvious one being that staff are required to have both the time and training to effectively develop these tools. In our two examples, the developers (first, second and third authors) were relatively proficient with R and have natural aptitudes for programming, but had their formal training in field-based population ecology. While some learning time was needed (a few weeks), the learning curve was relatively minor because there was no need to learn a new programming language (e.g. JavaScript, CSS, etc.). Moreover, a growing user base means that most programming issues were readily addressed through simple internet searches. We expect the accessibility and quality of these tools to continue to improve as more people in the fisheries community develop and use interactive documents. It is our experience that the upfront investment in time has already paid dividends in terms of efficiencies of delivering products for subsequent stock assessments, and in the quality of the data exploration and understanding. Scientific exploration is rarely a linear process, and the ability to navigate around a dashboard fits the natural way people pursue ideas.

>>>>

Northern cod (NAFO Divisions 2J3KL) has a rich history of tagging, starting in in 1954 [@taggart1995] and continuing to this day. The tagging and recovery data are captured in a standardized database, with fields typical of most tagging programs. This data base has over 600,000 records as of early 2019, with 2,000-10,000 tags deployed annually in recent decades. The tagging and capture data are used in the current assessment model for this stock [@cadigan2016], but tools to explore this extensive data set were limited, especially from a spatial perspective. To begin to explore and understand this large data set, we built a simple application using shiny and leaflet (Table \@ref(tab:pkg)) that allows a user to quickly and dynamically subset the data (e.g. ranges of release and capture years, specific geographic locations). As the application developed, it was incorporated into shinydashboard, to take advantage of the number of layout options available in shinydashboard (Table \@ref(tab:pkg)). Given the large number of tags released, often at nearby sites, visualizing the data with static mapping was particularly challenging. The markercluster (https://github.com/Leaflet/Leaflet.markercluster) function available in leaflet was particularly useful as a means to dynamically scale the level of pooling of spatial points (Figure \@ref(fig:map)). This basic mapping tool allowed us to quickly become familiar with the data, identify outliers and incorrect data entries, as well as explore options on how to spatially pool the data for subsequent demographic analysis. Further tabs were added to provide basic summaries of the selected data (Figure \@ref(fig:summary)).

@@

Synthesizing data from multiple sources presents a key challenge to stock assessment. Analyses of different data sources were traditionally carried out independently and the summaries or parameters from these analyses were used in the assessment model. This approach, however, is less than ideal because information may be lost and uncertainty may be unaccounted for when we "do statistics on the statistics" [@link1999; @maunder2013]. Such issues have largely been curtailed in contemporary stock assessments thanks to advances in statistical computing that have facilitated the analysis of all available data, in as raw a form as appropriate, in a single integrated analysis [@maunder2013]. Specifically, statistical modeling tools such as JAGS [@plummer2003], AD Model Builder [@fournier2012] and Template Model Builder [@kristensen2015] allow the construction of a joint likelihood for an array of observations to, in theory, extract as much information as possible about the biological and fishery processes. From a computational perspective, analyses of a variety of large data sets has never been easier. However, from a human perspective, contemporary stock assessment biologists are faced with the challenge of understanding and integrating data from multiple sources into a single model and communicating the methods and results to stakeholders and fisheries managers. This challenge was palpable for data-rich case of Northern cod.

The Northern cod stock off southern Labrador and eastern Newfoundland is one of the most well studied stocks in eastern Canada. As such, there are multiple monitoring programs that help inform the status of the stock and data from most of these programs have been integrated into a state-space stock assessment model, called NCAM [@cadigan2016]. The model includes information from research vessel autumn trawl surveys (1983-present), Sentinel fishery surveys (1995-present), inshore acoustic surveys (1995-2009), fishery catch-at-age compositions and partial fishery landings (1983-present), and tagging (1983-present). Using a series of observation equations, this TMB based model reduces thousands of historical data points into quantities such as recruitment, spawning stock biomass, fishing mortality and natural mortality. Once the model is fit to the data, the next step is to produce visual representations of the data and model output. The traditional approach would involve producing static presentations and documents with a series of figures and tables.  However, with large amounts of model inputs and outputs, this approach quickly becomes overwhelming for both the analyst and the stakeholders involved for at least two reasons. First, it is no longer feasible for the analyst to include and describe every figure and table produced in a single document. Second, it is difficult for stakeholders to efficiently digest the information that has been compressed into a series of static slides or pages. Interactive documents provide a potential solution to this problem as they allow much more information to be contained and accessible on a single screen.

In the pursuit of an easier and more efficient way to communicate results from NCAM, an interactive and self-contained "dashboard", called NCAM explorer, was developed for the 2018 assessment of Northern cod [@dwyer2019]. We used R-based packages (Table \@ref(tab:pkg)) to construct a tool for exploring the input and output of NCAM, specifically the flexdashboard package to group interactive plotly-based visuals into a dynamic document. We also used the crosstalk package to link the data displayed across multiple plots. Via R Markdown, the dashboard is rendered into a self-contained html file that is reproducible, interactive, and easy to update following modifications to the model or the addition of new data.

The NCAM dashboard (Supplement 2) contains a series of pages, the first of which provides terse point-form background on the model (page named "Background"). Subsequent pages provide a series of diagnostic plots for assessing model fits to catch ("Catch"), survey ("RV survey", "SN survey" and "SS survey"; accessed from the "Surveys" drop-down menu), and tagging ("Tagging") data. For instance, the "RV survey" page includes plots of observed and predicted values of mean numbers per tow captured in the research vessel survey (Figure \@ref(fig:rv)). The dashboard also includes pages focused on model estimates such as catchability and selectivity ("Catchability"), stock size and vital rates ("Trends"; Figure \@ref(fig:trends)), and stock productivity ("Productivity"). Finally, some results from a retrospective analyses are included under the "Retro" page, trends from different models are compared under the "Comps" page, details on the projections are accessed from the "Projections" drop-down menu ("Assumptions", "Past projections", "Retro projections", and "Results" pages), and key inputs and outputs are accessed from "Tables" drop-down menu ("Inputs", "Settings", "Outputs" pages). The plots and tables included in the dashboard are similar to those typically presented at assessment meetings and in research documents, however, there are two key benefits of this approach over the *modus operandi* of producing static documents and slides: 1) interactive plots nested in a dashboard permit relatively easy and efficient access to the details as it replaces scrolling through tens, if not hundreds, of pages or slides with mouse-clicks across pages holding data-rich illustrations (i.e. both broader patterns and finer details in the data are accessible via zooming and tooltips); and, 2) the automated nature of the dashboard circumvents the monotonous, time-consuming and error-prone task of copying and pasting figures, tables and values into documents and slides. Both benefits expedite the process of exploring a range of model configurations as the automated output facilitate quick views of standard diagnostics and the interactive plots facilitate detailed explorations and comparisons of models with different configurations. The dashboard also makes it easy to share the results with colleagues and stakeholders as it is rendered into a self-contained html file. This allows others to independently scrutinize details of both data and the model that are typically only accessible to the analyst. Such access improves the transparency of the stock assessment model which, in turn, leads to richer discussion and scrutiny of the biological and statistical rigor of the model. For instance, visualizations depicting the model assumptions, process errors and confidence intervals around the projections raised important questions on the impact of the assumptions on the projections from the model. These questions were raised during the first assessment meeting in which this tool was used and the ensuing discussion helped the assessment biologists, managers and fisheries representatives in the room gain a deeper understanding of the uncertainties and the patterns in the risk tables.

@@

The tools presented here are additions to the growing number of interactive stock assessment tools being developed and applied across the globe (see Appendix A). We believe that such tools represent an important a step forward in terms of open science and reproducible research in addition to being a major step forward in how stock assessments are presented and critiqued. Although these tools do not reveal the entire data pipeline, in conjunction with posting code and data on-line, they represent a significant advance over more traditional methods.   Similarly, these tools do not fully alleviate pragmatic issues associated with reproducible research.  For example, it is unlikely that an individual has the time to fully review and recreate a particular analysis in its entirety when most participants are already overcommitted [@banks2011].  Further, in the case of NCAM, considerable statistical experience and expertise is required simply to run the model. However, the dashboard approach removes many of these obstacles. The html file contains all model inputs, outputs and diagnostics in an open, and easy to use, interactive document that can be distributed prior to meetings.  As such, participants can assess the results and diagnostics at their leisure. Specific experience with the tools used to generate the results are not required to constructively and critically review of the results presented in the dashboard.

@@

We must acknowledge there are costs to adopting the use of interactive tools. The first and obvious one being that staff are required to have both the time and training to effectively develop these tools. In our two examples, the developers (first, second and third authors) were relatively proficient with R and have natural aptitudes for programming, but had their formal training in field-based population ecology. While some learning time was needed (a few weeks), the learning curve was relatively minor because there was no need to learn a new programming language (e.g. JavaScript, CSS, etc.). Moreover, a growing user base means that most programming issues were readily addressed through simple internet searches. We expect the accessibility and quality of these tools to continue to improve as more people in the fisheries community develop and use interactive documents. It is our experience that the upfront investment in time has already paid dividends in terms of efficiencies of delivering products for subsequent stock assessments, and in the quality of the data exploration and understanding. Finally, scientific exploration is rarely a linear process, and the ability to navigate around a dashboard fits the natural way people pursue ideas.
